---
title: "Using the US Dept. of Ed. College Scorecard API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{college-scorecard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
```

# The College Scorecard API

## Documentation

[This url](https://collegescorecard.ed.gov/data/documentation/) is the central site for documentation about this tool.

There are more detailed instructions at the tools [Github project](https://github.com/RTICWDT/open-data-maker/blob/master/API.md).

## Schools

These tools were developed to wrangle data about schools in the [Higher Education Regional Alliance (HERA)](https://herawisconsin.org). "HERA is a collaborative of 18 southeastern Wisconsin public and private 2- and 4-year colleges and universities, and partner institutions." [Higher Expectations for Racine County](https://www.higherexpectationsracinecounty.org) (HE), the creators of this package, is one of those institutions. The eighteen institutions, with some satellite campuses, can be seen in the following table:

```{r the-table-of-schools, fig.cap = "Schools in HERA"}
hera_schools <- readr::read_csv(
    system.file("extdata", "hera_schools.csv",
                package = "hiRx",
                mustWork = TRUE),
    show_col_types = FALSE
)

rlang::with_options(
    {
        knitr::kable(hera_schools)
    },
    knitr.kable.NA = ''
)

```

### Identifying schools and campuses

You can see from the table above that a single school may be associated with multiple units. These different units are identified by the `Unit ID` field. A unit may be a geographically distinct campus, as in the three Bryant & Stratton College locations in the data set. The unit might also capture the distinction between in-person and virtual course offerings, as seen in the "flex" options for UW-Milwaukee and UW-Parkside. The six-digit `OPE ID` field seems like the one to use for keeping track of multi-campus institutions that still count as single schools.

### Querying for the appropriate information

The College Report Card API has many different fields that you can query.
Our initial foray into using the API was for comparing graduation rates among years and schools.
The specific fields that we ended up needing are in the following table.

```{r api-fields}

api_fields <- readr::read_csv(
    system.file("extdata", "college reportcard api fields.csv",
                package = "hiRx"),
    show_col_types = FALSE
)

rlang::with_options(
    {
        knitr::kable(dplyr::select(api_fields, !.data$NOTES))
    },
    knitr.kable.NA = ''
)
```
