---
title: "laus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{laus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
content_size <- c(width = 7, height = 4)
content_dpi <- 96
knitr::opts_chunk$set(
    collapse = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>",
    cache = TRUE,
    dpi = content_dpi,
    fig.dim = content_size
)
```

```{r setup}
library(hiRx)
library(magrittr)
set.seed(42)

```

## LAUS Data

Local Area Unemployment Statistics are systems indicators of workforce health.
There are many ways to access these data, including a RESTful API and direct downloads of text files.
As of March 2022, downloading the entire data set in text form is almost as fast as an API query for just one complete series.
\U1f644
Consequently, this vignette is going to focus on fetching data from the static site.

### Online Layout

The root of the plain-text archive is https://download.bls.gov/pub/time.series/la.
Visiting that page yields an HTML table that documents all of the different files that one can fetch.
An excerpt of this table shows that it describes the files contained in the root.
The description includes a timestamp of when the file was last updated, the file's size in bytes, and the full name of the file.
The `bls_index_table` function also parses the file's name into an integer index and a short description of the file's contents.

```{r fetch-laus-index}
laus_index <- bls_index_table("pub/time.series/la")
laus_index %>%
    dplyr::slice(c(3, 5, 12, 18, 23, 45, 66, 80, 82)) %>%
    dplyr::arrange(.data$Index) %>%
    knitr::kable()
```

It turns out that the observations in these files are not unique.
If you just downloaded all the files, you would get many duplicate observations.
The files with the word "CurrentU" in their name contain data from every geography over a limited time span.
These data go back only to 1990.
The files that are organized geographically contain all of the records for their areas, going back to when they were first collected.
However, most of these data _also_ only go back to 1990.
The only areas with data from before then are states, census regions and divisions, New York City, and the LA area.
That means that we can fetch the older data for those regions' files and combine them with data for everywhere else from the current data files.

### Downloading

The next several blocks of code demonstrate how to fetch ALL of the LAUS data.
These downloads involve many megabytes, so the code is not actually run.

It takes three steps to assemble all of the LAUS observations into one tibble.

First, create a "laus_current" tibble by downloading data sets with "Current" in their name.
The `laus_fetch_observations` function uses the `laus_index` file to fetch the data.

This takes about 100 seconds.

```{r fetch-current-laus, eval = FALSE}
laus_current <- laus_fetch_observations(index_table = laus_index,
                                        pattern = "CurrentU")
```

Second, create a "laus_legacy" tibble.
We need NYC and the LA are, so we'll download the data sets with "NewYork" and "California" in their names, then filter out all of the state-wide values.
Then we'll bind that result to the data sets with "AllStates" or "Region" in their names.
We only need the pre-1990 data in these tables, so we'll filter them out here.
The lack of spaces in "AllStates" and "NewYork" is not a mistake.
This takes about 10 seconds.

```{r fetch-legacy-laus, eval = FALSE}
laus_legacy <- laus_index %>%
    laus_fetch_observations(pattern = "California|NewYork") %>%
    dplyr::filter(!stringr::str_detect(.data$series_id, "^.{3}ST")) %>%
    dplyr::bind_rows(laus_fetch_observations(laus_index,
                                             "Region|AllStates")) %>%
    dplyr::filter(.data$year < 1990)
```

Third, bind the rows together into a "laus_observations" tibble and remove the tibbles from the first two steps.
This only takes one or two seconds.

```{r laus-assemble-observations, eval = FALSE}
laus_observations <- dplyr::bind_rows(laus_legacy,
                                      laus_current)
rm(laus_legacy, laus_current)
```

As of May 2022, this `laus_observations` tibble has 13,727,807 observations.

Since the result from [Downloading](#downloading) has more than 10 million rows, it isn't suitable for use in a vignette as example data.
Instead, I will use a smaller example with "only" 6704 observations.
As you can see below, it has sixteen different series.

```{r show-laus-example}
laus_example <- NULL
data("laus_example", package = "hiRx")
laus_example %>%
    dplyr::count(.data$series_id) %>%
    knitr::kable()
```

### Wrangling

The series IDs by themselves are not particularly helpful.
They contain four pieces of information packed together:

- The type of BLS data set that the series belongs to (which is always "LA" for LAUS data.)
- Whether or not the observations are seasonally adjusted ("S" for yes, "U" for no).
- The geographic location that the series refers to, itself a 15-character composite of
  - A 2-character code describing the kind of geographic area.
  - A 2-digit FIPS code for the state, census region, or census division the area belongs to.
  - A 0- to 11-digit FIPS code for a specific area within a state
- What kind of observation the series reports on (e.g. "03" for Unemployment Rate)

#### Metadata

The BLS site has [a table that provides documentation](https://download.bls.gov/pub/time.series/la/la.series) for every one of its series.
Here is a distilled version with just the useful metadata about the series in the example data.

```{r laus-example-series-metadata}
laus_series_metadata <- laus_fetch_table(
    "la.series",
    readr::cols(
        series_id = readr::col_character(),
        area_type_code = readr::col_character(),
        area_code = readr::col_character(),
        measure_code = readr::col_integer(),
        seasonal = readr::col_character(),
        srd_code = readr::col_integer(),
        series_title = readr::col_character(),
        footnote_codes = readr::col_character(),
        begin_year = readr::col_integer(),
        begin_period = readr::col_character(),
        end_year = readr::col_integer(),
        end_period = readr::col_character()
    )
)

laus_example_metadata <- laus_example %>%
    dplyr::distinct(
        .data$series_id
    ) %>%
    dplyr::left_join(
        laus_series_metadata,
        by = "series_id"
    ) %>%
    dplyr::select(
        where(~ dplyr::n_distinct(.) > 1)
    )

laus_example_metadata %>%
    knitr::kable(
        col.names = names(.) %>%
            stringr::str_replace_all("_", " ") %>%
            stringr::str_to_title()
    )
```

You can see that the series title is a combination of the names of its area and its measure.
BLS also provides a [table of area names for each area code](https://download.bls.gov/pub/time.series/la/la.areamaps).
The entries for our four example areas look like this:

```{r laus-example-area-names}
laus_areas <- laus_fetch_table(
    "la.areamaps",
    readr::cols(
        area_type_code = readr::col_character(),
        area_code = readr::col_character(),
        area_text = readr::col_character(),
        display_level = readr::col_integer(),
        selectable = readr::col_logical(),
        sort_sequence = readr::col_integer()
    )
)

laus_example_areas <- laus_example_metadata %>%
    dplyr::distinct(.data$area_code) %>%
    dplyr::left_join(laus_areas, by = "area_code") %>%
    dplyr::select(where(~dplyr::n_distinct(.) > 1))

laus_example_areas %>%
    knitr::kable(
        col.names = names(.) %>%
            stringr::str_replace_all("_", " ") %>%
            stringr::str_to_title()
    )
```

The descriptions of the measures are in their own [table on the BLS site](https://download.bls.gov/pub/time.series/la/la.measure).
The example data only contain the first four types of measures that are found in the following table.

```{r laus-example-measure-glossary}
laus_glossary_measure <- laus_fetch_table(
    "la.measure",
    readr::cols(
        measure_code = readr::col_integer(),
        measure_text = readr::col_character()
    )
)

laus_glossary_measure %>%
    knitr::kable(
        col.names = names(.) %>%
            stringr::str_replace_all("_", " ") %>%
            stringr::str_to_title()
    )
```

There are several such key-value files on the LAUS site.
I'm going to fold them all into a single tibble called `laus_glossaries`.

```{r fetch-laus-glossaries, include = FALSE}
laus_glossaries <- c(
    "area_type",
    "footnote",
    "measure",
    "period",
    "seasonal",
    "state_region_division"
) %>%
    rlang::set_names() %>%
    purrr::map(
        ~ laus_fetch_table(paste0("la.", .),
                           readr::cols(
                               .default = readr::col_character()
                           ))
    ) %>%
    purrr::imap(
        ~ .x %>%
            tidyr::pivot_longer(
                cols = 2:last_col()
            ) %>%
            dplyr::rename(
                key = 1
            ) %>%
            dplyr::mutate(
                table = .y,
                .before = .data$key
            )
    ) %>%
    dplyr::bind_rows()
```

### Plotting

So, to look at trends in unemployment rates among the four example areas:

```{r laus-example-plot-unemployment-rates}
laus_example_metadata %>%
    dplyr::filter(
        .data$measure_code == 3
    ) %>%
    dplyr::left_join(
        laus_example_areas,
        by = "area_code"
    ) %>%
    dplyr::left_join(
        laus_example %>%
            dplyr::filter(.data$period == "M13"),
        by = "series_id"
    ) %>%
    ggplot2::ggplot(ggplot2::aes(
        x = .data$year,
        y = .data$value,
        color = stringr::str_extract(
            .data$area_text,
            "^\\w+")
    )) + 
    ggplot2::geom_line() + 
    ggplot2::labs(x = "Year",
                  y = "Unemployment",
                  color = NULL) +
    scale_color_higher_ex("tableau") +
    ggplot2::scale_y_continuous(
        labels = scales::label_percent(scale = 1,
                                       accuracy = 1),
        limits = c(0, NA)
    ) + 
    ggplot2::theme_minimal()
```
