---
title: "Merging Census Areas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging Census Areas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(magrittr)

.CONTENT_SIZE <- c(width = 7L,
                   height = 4L)
.CONTENT_DPI <- 200L

rlang::exec(knitr::opts_chunk$set,
            !!!hiRx::CHUNK_OPTIONS_FOR_KNITR,
            dpi = .CONTENT_DPI,
            fig.dim = .CONTENT_SIZE
)

DATA_PATH <- Sys.getenv("DATA_PATH")

SHAPEFILE_PATH <- Sys.getenv("SHAPEFILE_PATH")

east_of_I94 <- function(.x) {
    dplyr::filter(.x,
                  .data$INTPTLON > -87.96)
}

west_of_lake_michigan <- function(.x) {
    dplyr::filter(.x,
                  .data$INTPTLON < -87.7)
}

UTM_ZONE <- sf::st_crs(32616L)

```

## Data Items

There are two kinds of data to collect. Shapefile data define the locations of geographic areas. Census table data report on demographic information within those areas.

### Geographic Data

It is possible to acquire county-level shapefiles for the 2000 and 2010 censuses. The 2020 census data are only provided as statewide files, so that file is filtered by county. The full list of Racine County subdivisions contains many places that we don't work with. The features must be filtered to include only land areas and only areas east of Interstate 94 and west of Lake Michigan.

**change** we are now using the whole county

#### County Subdivisions

The simple features that describe Higher Expectations' partner municipalities are found in TIGER/Line's "county subdivisions" shapefiles. 

```{r load_places}
PLACE_FILES <- c(
    `2000` = "tl_2010_55101_cousub00",
    `2010` = "tl_2010_55101_cousub10",
    `2020` = "tl_2020_55_cousub"
)

.MUNICIPALITIES <- tibble::tribble(
    ~ NAMELSAD,                        ~ Municipality,
    "County subdivisions not defined", "Lake Michigan",
    "Burlington city",                 "Burlington",
    "Burlington town",                 "Unincorporated",
    "Caledonia town",                  "Caledonia",
    "Caledonia village",               "Caledonia",
    "Dover town",                      "Unincorporated",
    "Elmwood Park village",            "Elmwood Park",
    "Mount Pleasant town",             "Mount Pleasant",
    "Mount Pleasant village",          "Mount Pleasant",
    "North Bay village",               "North Bay",
    "Norway town",                     "Unincorporated",
    "Racine city",                     "Racine",
    "Raymond town",                    "Raymond",
    "Raymond village",                 "Raymond",
    "Rochester town",                  "Rochester",
    "Rochester village",               "Rochester",
    "Sturtevant village",              "Sturtevant",
    "Union Grove village",             "Union Grove",
    "Waterford town",                  "Unincorporated",
    "Waterford village",               "Waterford",
    "Wind Point village",              "Wind Point",
    "Yorkville town",                  "Yorkville",
    "Yorkville village",               "Yorkville"
)

.REGIONS <- tibble::tribble(
    ~ Municipality,   ~ Region,
    "Burlington",     "West of the I",
    "Caledonia",      "Suburban RUSD",
    "Elmwood Park",   "Suburban RUSD",
    "Lake Michigan",  NA,
    "Mount Pleasant", "Suburban RUSD",
    "North Bay",      "Suburban RUSD",
    "Racine",         "City of Racine",
    "Raymond",        "West of the I",
    "Rochester",      "West of the I",
    "Sturtevant",     "Suburban RUSD",
    "Unincorporated", "West of the I",
    "Union Grove",    "West of the I",
    "Waterford",      "West of the I",
    "Wind Point",     "Suburban RUSD",
    "Yorkville",      "West of the I"
)

PLACES <- PLACE_FILES |>
    hercgeos::batch_read_tigershapes(
        .names_to = "File"
    ) |>
    dplyr::filter(
        SHAPEFILE_PATH,
        .data$COUNTYFP == 101L,
        .data$INTPTLON < -87.7
    )

MUNIS <- PLACES |>
    dplyr::left_join(
        .MUNICIPALITIES,
        by = "NAMELSAD"
    ) |>
    dplyr::group_by(
        .data$Year,
        .data$Municipality
    ) |>
    dplyr::summarize(
        dplyr::across(
            tidyselect::all_of(c("ALAND", "AWATER")),
            ~ sum(., na.rm = TRUE)
        ),
        dplyr::across(
            tidyselect::starts_with("INTPTL"),
            ~ median(., na.rm = TRUE)
        ),
        Area = (.data$ALAND + .data$AWATER) / 1e6,
        geometry = sf::st_union(.data$geometry),
        .groups = "keep"
    ) |>
    dplyr::ungroup() |>
    dplyr::left_join(
        .REGIONS,
        by = "Municipality"
    )
```

#### Census Blocks

The decennial censuses report data down to the census block. The population within each block will be used to determine what proportion of larger areas, like block groups or tracts, belong to each municipality. These larger areas are the finest level of geographic specificity that ACS data get to.

```{r load_blocks}
BLOCK_FILES <- c(
    `2000` = "tl_2010_55101_tabblock00",
    `2010` = "tl_2010_55101_tabblock10",
    `2020` = "tl_2020_55_tabblock20"
)

BLOCKS <- BLOCK_FILES |>
    hiRx::census_tigershapes_combine(
        SHAPEFILE_PATH,
        .data$COUNTYFP == 101L
    )
```

#### Assign census blocks to the places that contain them.

```{r assign_blocks_to_munis}

MUNI_BLOCKS <- MUNIS |>
        sf::st_transform(UTM_ZONE) |>
        tidyr::nest(
            locale = !tidyselect::all_of(c(
                "Region",
                "Municipality",
                "Year"
                ))
        ) |>
        dplyr::mutate(
            blocks = purrr::map2(
                .data$locale,
                .data$Year,
                ~ dplyr::filter(BLOCKS,
                                .data$Year == .y) |>
                    sf::st_transform(UTM_ZONE) |>
                    hiRx::geom_filter_by_interior_intersection(.x) |>
                    dplyr::select(tidyselect::all_of(c("TRACTCE",
                                                       "BLOCKCE"))) |>
                    sf::st_drop_geometry()
            )
        ) |>
        dplyr::select(
            !.data$locale
        ) |>
        tidyr::unnest(
            cols = .data$blocks
        )

```

#### Census Tracts

Using a union command from the simple features library, we can assemble the individual block geometries into the tracts that contain them. This is also an opportunity to compute summary features, such as the population within each census tract and the area of each tract in square kilometers.

```{r aggregate_tracts}
TRACTS <- BLOCKS |>
    dplyr::group_by(
        .data$Year,
        .data$TRACTCE
    ) |>
    dplyr::summarize(
        Area = sum(.data$ALAND + .data$AWATER) / 1e6,
        geometry = sf::st_union(.data$geometry),
        .groups = "keep"
    ) |>
    dplyr::ungroup()
```

### Demographic Data

#### Decennial Census Tables

The decennial censuses are the only ones that have data at the block level of geographic detail. It turns out that the easiest way to access these data is ask for all of the blocks from each census.
Table formats and contents vary across decennial censuses, which impacts the API calls.

##### Population Within Census Blocks

```{r decennial_population_settings}
.DECENNIAL_POPULATION_VARIABLES <- tibble::tribble(
    ~ year, ~ Variable, ~ Race,
    
    2000L, "PL002002", "Latin@",
    2000L, "PL002003", "All Other Races",
    2000L, "PL002005", "White",
    2000L, "PL002006", "Black",

    2010L, "P002002", "Latin@",
    2010L, "P002003", "All Other Races",
    2010L, "P002005", "White",
    2010L, "P002006", "Black",

    2020L, "P2_002N", "Latin@",
    2020L, "P2_003N", "All Other Races",
    2020L, "P2_005N", "White",
    2020L, "P2_006N", "Black",
)

.DECENNIAL_POPULATION_SETTINGS <- .DECENNIAL_POPULATION_VARIABLES |>
    dplyr::select(
        .data$year,
        .data$Variable
        ) |>
    tidyr::nest(
        data = .data$Variable
        ) |>
    dplyr::mutate(
        variables = purrr::map_chr(data,
                                   ~ paste0(.$Variable, collapse = ",")),
        .keep = "unused")

```

The census API returns JSON data in a 2D, list-of-lists format. The following chunk makes a total of 3 api calls and retrieves thousands of observations (one for each block in Racine County).

```{r acquire_populations}
RAW_POPS <- .DECENNIAL_POPULATION_SETTINGS |>
    purrr::pmap_chr(hiRx::decennial_query_url,
                    summary_table = "pl",
                    for_geo = "block",
                    for_items = "*",
                    state = 55,
                    county = 101
    ) |>
    magrittr::set_names(., .DECENNIAL_POPULATION_SETTINGS$year) |>
    purrr::map(hiRx::census_do_api_query)
```

There are some formatting problems in the data from the 2000 census. For example, the tract field is usually 6 digits long, so some responses need extra zeros on the right side. This formatting problem makes it really difficult to query on a per-tract basis. The total population variable also has a different name in some decennial census. Each of those names got its own column after the `bind rows` step above. One column has numbers for each census, and the rest are `NA`s. Here, those columns are stacked into a single one, and the `NA` variables are expunged.

```{r wrangle_populations}
POPS <- RAW_POPS |>
    purrr::map(
        hiRx::census_json_to_tibble,
        variable_names = .DECENNIAL_POPULATION_VARIABLES$Variable
    ) |>
    dplyr::bind_rows(
        .id = "Year"
    ) |>
    dplyr::mutate(
        Year = as.integer(.data$Year),
        tract = stringr::str_pad(
            .data$tract,
            width = 6,
            side = "right",
            pad = "0"
        )
    ) |>
    dplyr::left_join(
        .DECENNIAL_POPULATION_VARIABLES,
        by = c(Year = "year", "Variable")
    ) |>
    dplyr::select(
        !.data$Variable
    )
```

Higher Expectations consolidates the Census's Race and Ethnicity classification in a pretty typical way, carrying on the ugly "one-drop" legacy of our country.
Think of it as the "racist cab-driver" classification system.

"Black or African American" gets classified as "Black," regardless of Latin@ ethnicity.
"White" gets classified as "White" when its ethnicity is not Latin@.
"All Other Races" and "White" get classified as "Latin@" when that is their ethnicity.
"All Other Races" is everything that did not get classified as "Black", "Latin@", or "White".

```{r wrangle_races}

RACES <- POPS |>
    tidyr::pivot_wider(
        names_from = .data$Race,
        values_from = .data$Population
    ) |>
    dplyr::mutate(
        `All Other Races` = .data$`All Other Races` - .data$Black - .data$White
    ) |>
    tidyr::pivot_longer(
        cols = tidyselect::any_of(hiRx::RACE_LABELS_RACINE),
        names_to = "Race/Ethnicity",
        values_to = "Population"
    ) |>
    dplyr::mutate(
        `Race/Ethnicity` = factor(.data$`Race/Ethnicity`,
                                  levels = hiRx::RACE_LABELS_RACINE)
    )
```

#### Populations within Each Tract

There is no need to perform any spatial analysis to find the population totals within each tract.
Grouping and summarizing by geographic factors is sufficient.

```{r tract_races}
TRACT_RACES <- RACES |>
    dplyr::group_by(
        .data$Year,
        .data$state,
        .data$county,
        .data$tract,
        .data$`Race/Ethnicity`
    ) |>
    dplyr::summarize(
        `Tract Population` = sum(.data$Population),
        .groups = "keep"
    ) |>
    dplyr::ungroup()
```

### Combining Geography and Demography

The most useful combination of geography and demography is the population of each race within each combination of tract and place.
Many tracts occur entirely within one county subdivision, but others overlap subdivision borders.
Keeping track of the population within each intersection makes it possible to compute weighted statistics for each municipality for data that are collected at the tract level.

```{r muni_tract_races}
MUNI_TRACT_RACES <- RACES |>
    dplyr::inner_join(
        MUNI_BLOCKS,
        by = c("Year", tract = "TRACTCE", block = "BLOCKCE")
    ) |>
    dplyr::group_by(
        .data$Year,
        .data$Region,
        .data$Municipality,
        .data$tract,
        .data$`Race/Ethnicity`,
    ) |>
    dplyr::summarize(
        `Municipal Population` = sum(.data$Population),
        .groups = "keep"
    ) |>
    dplyr::ungroup() |>
    dplyr::inner_join(
        TRACT_RACES,
        by = c("Year", "tract", "Race/Ethnicity")
    )
```

#### Place Populations Based on Block Populations

```{r muni_pops}
.block_fields <- c("Year", "tract", "block")

MUNI_POPULATIONS <- MUNI_BLOCKS |>
    dplyr::rename(
        tract = "TRACTCE",
        block = "BLOCKCE"
    ) |>
    dplyr::left_join(RACES,
                     by = .block_fields
    ) |>
    dplyr::group_by(
        .data$Year,
        .data$Region,
        .data$Municipality,
        .data$`Race/Ethnicity`
    ) |>
    dplyr::summarize(
        Population = sum(.data$Population),
        .groups = "keep"
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
        Municipality = forcats::fct_reorder(.data$Municipality,
                                            .data$Population)
    )
```

```{r table_of_muni_pops_per_year}
MUNI_POPULATIONS |>
    dplyr::group_by(
        .data$Region,
        .data$Municipality,
        .data$Year,
    ) |>
    dplyr::summarize(
        Population = sum(.data$Population),
        .groups = "keep"
    ) |>
    tidyr::pivot_wider(
        names_from = .data$Year,
        values_from = .data$Population
    ) |>
    dplyr::arrange(
        .data$Region,
        dplyr::desc(.data$`2020`)
    ) |>
    knitr::kable(
        format.args = list(big.mark = ",")
    )
```

```{r area_plots_of_region_race_populations}

MUNI_POPULATIONS |>
    dplyr::filter(
        !is.na(.data$Region)
    ) |>
    dplyr::group_by(
        .data$Year,
        .data$Region,
        .data$`Race/Ethnicity`
    ) |>
    dplyr::summarize(
        Population = sum(.data$Population,
                         na.rm = TRUE),
        .groups = "keep"
    ) |>
    dplyr::ungroup() |>
    ggplot2::ggplot(ggplot2::aes(
        x = .data$Year,
        y = .data$Population,
        fill = .data$`Race/Ethnicity`
    )) +
    ggplot2::geom_area() +
    ggplot2::scale_y_continuous(
        name = "Population",
        labels = scales::label_comma(accuracy = 1,
                                     scale = 1/1000,
                                     suffix = "K")
    ) +
    hiRx::scale_fill_higher_ex(
        "race",
        n.colors = 4
    ) +
    ggplot2::facet_wrap(ggplot2::vars(
        .data$Region
    )) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
        legend.position = "top"
    )

```

```{r bar_plots_for_munis}
PLOTS_FOR_MUNIS <- MUNI_POPULATIONS |>
    tidyr::nest(data = !.data$Municipality) |>
    dplyr::pull(.data$data,
                name = .data$Municipality) |>
    purrr::imap(~ .x |>
                    ggplot2::ggplot(
                        ggplot2::aes(x = .data$Year,
                                     y = .data$Population,
                                     fill = .data$`Race/Ethnicity`)
                    ) +
                    ggplot2::scale_x_continuous(
                        breaks = c(2000, 2010, 2020),
                        minor_breaks = NULL
                    ) +
                    ggplot2::scale_y_continuous(
                        labels = scales::label_comma(),
                        minor_breaks = NULL,
                        limits = c(0, NA)
                    ) +
                    hiRx::scale_fill_higher_ex("race", n.colors = 4) +
                    ggplot2::labs(title = .y) +
                    ggplot2::theme_minimal() +
                    ggplot2::theme(legend.position = "bottom")
    )

BAR_PLOTS_FOR_MUNIS <- PLOTS_FOR_MUNIS |>
    purrr::map( ~ . + 
            ggplot2::geom_col(position = "dodge",
                              col = hiRx::higher_ex_cols("gray"))
    )

purrr::walk(BAR_PLOTS_FOR_MUNIS, print)
```

```{r area_plots_for_munis}
AREA_PLOTS_FOR_MUNIS <- PLOTS_FOR_MUNIS |>
    purrr::map( ~ . + 
            ggplot2::geom_area(col = hiRx::higher_ex_cols("gray"))
    )

purrr::walk(AREA_PLOTS_FOR_MUNIS, print)

```

#### Tract Populations by Race and Municipality

```{r population_by_place_and_tract}
MUNI_TRACT_POPS <- MUNI_BLOCKS |>
    dplyr::select(
        .data$Municipality,
        .data$Year,
        tract = .data$TRACTCE,
        block = .data$BLOCKCE
        ) |>
    dplyr::right_join(
        RACES,
        by = c("Year", "tract",  "block")
        ) |>
    dplyr::group_by(
        .data$Year,
        .data$Municipality,
        .data$tract,
        .data$`Race/Ethnicity`
        ) |>
    dplyr::summarize(
        Population = sum(.data$Population),
        .groups = "keep"
        ) |>
    dplyr::ungroup()
```

## Racine-Specific Examples of Some Patterns

Comparing Racine's boundaries to those of the tracts finds the tracts that make up the city.

```{r find_tracts_in_racine}

RACINE_CITY <- MUNIS |>
    dplyr::filter(
        stringr::str_detect(.data$Municipality, "Racine")
    ) |>
    dplyr::select(
        where(~ dplyr::n_distinct(.) > 1)
    ) |>
    sf::st_transform(UTM_ZONE)

RACINE_TRACTS <- TRACTS |>
        sf::st_transform(UTM_ZONE) |>
    tidyr::nest(
        small = !.data$Year
    ) |>
    dplyr::mutate(
        racine = purrr::map2(
            .data$small,
            .data$Year,
            ~ hiRx::geom_filter_by_interior_intersection(
                .x,
                dplyr::filter(RACINE_CITY,
                              .data$Year == .y)
            )
        )
    ) |>
    dplyr::select(
        !.data$small
    ) |>
    tidyr::unnest(
        cols = .data$racine
    ) |>
    sf::st_as_sf()

ggplot2::ggplot(RACINE_TRACTS,
                ggplot2::aes(fill = .data$TRACTCE)) +
    ggplot2::geom_sf(show.legend = FALSE,
                     col = "white") +
    ggplot2::geom_sf(data = RACINE_CITY,
                     col = "black",
                     fill = NA) +
    ggplot2::scale_x_continuous(
        breaks = seq(-87.9, -87.75, by = 0.05)
    ) +
    ggplot2::scale_y_continuous(
        breaks = seq(42.65, 42.85, by = 0.05)
    ) +
    ggplot2::facet_wrap(ggplot2::vars(.data$Year)) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
        legend.position = "bottom",
        strip.text = ggplot2::element_text(size = ggplot2::rel(3/2))
    )

```

#### Population Density of Racine's Census Tracts

```{r plot_racine_tracts_by_pop_density}
RACINE_TRACTS |>
    dplyr::left_join(
        MUNI_TRACT_POPS |>
            dplyr::filter(
                stringr::str_detect(.data$Municipality,
                                    "Racine")
            ) |>
            dplyr::group_by(
                .data$Year,
                .data$tract
            ) |>
            dplyr::summarize(
                Population = sum(.data$Population),
                .groups = "keep"
            ) |>
            dplyr::ungroup(),
        by = c("Year", TRACTCE = "tract")
    ) |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(
        ggplot2::aes(fill = .data$Population / .data$Area)
    ) +
    ggplot2::geom_sf(
        data = RACINE_CITY,
        col = "red",
        fill = NA
    ) +
    ggplot2::facet_wrap(
        ggplot2::vars(.data$Year)
    ) +
    ggplot2::scale_x_continuous(
        breaks = seq(-87.9, -87.75, by = 0.05)
    ) +
    ggplot2::scale_y_continuous(
        breaks = seq(42.65, 42.85, by = 0.05)
    ) +
    ggplot2::scale_fill_continuous(
        limits = c(0, 4000),
        breaks = scales::breaks_width(2000),
        labels = scales::label_comma()
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
        legend.position = "bottom",
        strip.text = ggplot2::element_text(size = ggplot2::rel(3/2))
    ) +
    ggplot2::labs(
        subtitle = "The City of Racine's borders are in red.",
        fill = "People per Sq. Km."
    )
```

#### Population Density of Racine's Census Tracts, by Race

```{r plot_racine_tracts_by_race_pop_density, fig.dim=c(10, 10)}
RACINE_TRACTS |>
    dplyr::left_join(
        MUNI_TRACT_POPS,
        by = c("Year", TRACTCE = "tract")
    ) |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(
        ggplot2::aes(fill = .data$Population)
    ) +
    ggplot2::geom_sf(
        data = RACINE_CITY,
        col = "red",
        fill = NA
    ) +
    ggplot2::scale_x_continuous(
        breaks = seq(-87.9, -87.75, by = 0.05)
    ) +
    ggplot2::scale_y_continuous(
        breaks = seq(42.65, 42.85, by = 0.05)
    ) +
    ggplot2::scale_fill_continuous(
        name = "People",
        labels = scales::label_comma(accuracy = 1,
                                     scale = 1/1000,
                                     suffix = "K")
    ) +
    ggplot2::facet_grid(
        rows = ggplot2::vars(.data$Year),
        cols = ggplot2::vars(.data$`Race/Ethnicity`)
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
        legend.position = "bottom",
        strip.text = ggplot2::element_text(size = ggplot2::rel(3/2))
    ) +
    ggplot2::labs(
        subtitle = "The City of Racine's borders are in red."
    )
```
