---
title: "Neighborhood-Atlas-Area-Deprivation-Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neighborhood-Atlas-Area-Deprivation-Index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}

library(magrittr)

rlang::exec(knitr::opts_chunk$set,
            hiRx::CHUNK_OPTIONS_FOR_KNITR)

.WHOLE_US <- "The Rest of the Country"
.STATE_NAME <- "Wisconsin"
.COUNTY_NAME <- "Racine County"
.STATE_FIPS <- 55L
.COUNTY_FIPS <- 101L


```

## Data Acquisition

### Neighborhood Atlas of Area Deprivation

The Atlas' authors prefer that we log in and download the data from them.
I'm sure that helps them with their metrics for grant compliance, so you should do that!
Go to the Neighborhood Atlas [website][neighborhood-atlas].
Create an account.
Download the national Block-Group level Deprivation Index data.
Change the `.DATA_FILE` statement below to point to your data file.

```{r adi_data_file}
.DATA_FILE <- hiRx::input_path("Downloads",
                              "US_2020_ADI_Census Block Group_v3.2.csv")

```

### Shapefiles of Census Block Groups

I acquired my shapefiles from the US Census Tiger/Line service.
[This link][tigerline] should take you right to a form where you can choose to download all of the block group shapes for one state.
I am in Wisconsin and use that.
You could (should?) pick your own state.
You'll need to unzip the contents, keep them together, and point R at the "shp" file.

```{r tigerline_data_file}

.SHAPEFILE <- Sys.getenv("SHAPEFILE_PATH") %>%
    file.path(
        "2020",
        "tl_2020_55_bg.shp"
    )

```

## Data Wrangling

### Data Format

The file has five fields, and each line describes one census block group.
The first is a sequential index that seems safe to ignore.
The second is the FIPS-based code for joining these data to a shapefile from the [National Historical Geographic Information System][nhgis].
The third is the national percentile ranking of the area deprivation index.
The fourth is the within-state decile ranking of the area deprivation index.
The fifth is the GEOID code of the block group.

```{r file_format_definitions}

.FIELD_TYPES <- readr::cols(
    OBJECTID = readr::col_skip(),
    GISJOIN = readr::col_character(),
    ADI_NATRANK = readr::col_integer(),
    ADI_STATERNK = readr::col_integer(),
    FIPS = readr::col_character()
)

```

Both the `ADI_NATRANK` and `ADI_STATERANK` fields contain missing values.
These data are not necessarily missing, but have been suppressed for privacy or data quality reasons.
I'm not entirely sure what the codes mean, although the authors do give at least a [partial explanation][na-explanation].
for blocks when the data from the census block group needed to be suppressed to protect the privacy of its residents.
The item names in the array below make a guess here about what the codes mean, but it doesn't really matter what they are because I am going to convert them all to `NA`s anyway.

```{r suppressed_data_definitions}

.SUPPRESSED_VALUES <-  c(
    `Group Quarters` = "GQ",
    `Too Few Housing Units` = "PH",
    `Both` = "GQ-PH",
    `Fewer than 100 People` = "QDI"
)

```

The GEOID code is a string that contains four integers that are also FIPS codes.
The US Census Bureau has a good [explanation][census-fips-explanation] of the layout.
Each of the four component codes is padded on the left with zeros to maintain consistent spacing.
The first two digits are the block groups' state FIPS codes.
The next three digits are county.
The next six are the census tract.
The final digit is the block group.

```{r geoid_code_definitions}

.GEOID_PARTS <- c(
    "State",
    "County",
    "Tract",
    "Block Group"
)

.GEOID_BREAKS <- cumsum(c(
    2L,
    3L,
    6L
))

```

### Actually Read the ADI Data

Now that we have defined all of the parameters for importing and wrangling the data, we can do so.
Note that we are removing the block groups with suppressed data.

```{r read_adi_data}

ADI_BY_BG <- .DATA_FILE %>%
    readr::read_csv(
        file = .,
        col_types = .FIELD_TYPES,
        na = .SUPPRESSED_VALUES
    ) %>%
    dplyr::filter(
        !is.na(.data$ADI_NATRANK),
        !is.na(.data$ADI_STATERNK)
    ) %>%
    dplyr::rename(
        GEOID = "FIPS"
    ) %>%
    tidyr::separate(
        col = .data$GEOID,
        into = .GEOID_PARTS,
        sep = .GEOID_BREAKS,
        convert = TRUE,
        remove = FALSE
    )

```

### Load Census Block Group Map Data

Next, let's read in some shapefiles so we can make a map.
We'll mutate the data so that the tacky character-typed FIPS fields are wholesome and useful integers.
We'll also make some of the column names more human-readable.

```{r import_shapefiles}

SIMPLE_FEATURES_BY_BG <- .SHAPEFILE %>%
    sf::read_sf() %>%
    dplyr::filter(
        .data$AWATER < 1e8
    ) %>%
    dplyr::mutate(
        dplyr::across(tidyselect::all_of(c("STATEFP",
                                           "COUNTYFP",
                                           "TRACTCE",
                                           "BLKGRPCE")),
                      readr::parse_integer)
    ) %>%
    dplyr::rename(
        State = "STATEFP",
        County = "COUNTYFP",
        Tract = "TRACTCE",
        `Block Group` = "BLKGRPCE",
    )
```

### Filter and Join the ADI and Shapefile Data

We can't plot the whole country, that would take to long.
Let's pick a US County at random ... Racine County, WI!
Coincidentally, I just happen to know that Wisconsin's FIPS code is 55 and Racine County's is 101.
Feel free to substitute your own state and county codes

```{r just_one_county}

COUNTY <- dplyr::left_join(
    dplyr::filter(SIMPLE_FEATURES_BY_BG,
                  .data$State == .STATE_FIPS,
                  .data$County == .COUNTY_FIPS),
    dplyr::filter(ADI_BY_BG,
                  .data$State == .STATE_FIPS,
                  .data$County == .COUNTY_FIPS),
    by = c("State",
           "County",
           "Tract",
           "Block Group",
           "GEOID")
)

```

## Visualize

There are some common settings that we're going to use in several plots.
We will focus on each block group's national percentile position in the Area Deprivation Index.
It should be more sensitive to local conditions because it is finer-grained than the state decile.
Unless, perhaps, you live in Wyoming or Delaware.

```{r settings_for_graphing_adi_natrank}

generic_settings <- list(
    ggplot2::theme_minimal(),
    ggplot2::theme(
        legend.position = "top"
    )
)

national_rank_axis = list(
    ggplot2::scale_y_continuous(
        name = "National Percentile of ADI",
        labels = scales::label_percent(accuracy = 1,
                                       scale = 1),
        limits = c(0, 100)
    )
)

national_rank_gradient_settings <- list(
    limits = c(0, 100),
    midpoint = 50,
    low = hiRx::higher_ex_cols("Medium Blue"),
    mid = ggplot2::alpha(hiRx::higher_ex_cols("Gray"),
                         alpha = 0.2),
    high = hiRx::higher_ex_cols("Gold"),
    na.value = ggplot2::alpha(hiRx::higher_ex_cols("Tableau Mauve"),
                              alpha = 0.1)
)

```

### A Sanity-check Scatterplot of National Data

The two quantile fields, National Percentile and State Decile, will obviously be correlated.
We'll plot them anyway just to highlight that we will be focusing on our county within our state.

```{r quantile_plot}

ADI_BY_BG %>%
    dplyr::mutate(
        Color = factor(
            dplyr::case_when(
                .data$State == .STATE_FIPS &
                    .data$County == .COUNTY_FIPS
                ~ .COUNTY_NAME,
                .data$State == .STATE_FIPS ~ .STATE_NAME,
                TRUE ~ .WHOLE_US
            ),
            levels = c(.WHOLE_US, .STATE_NAME, .COUNTY_NAME)
        )
    ) %>%
    dplyr::arrange(
        .data$Color
    ) %>%
    ggplot2::ggplot(ggplot2::aes(
        x = .data$ADI_STATERNK,
        y = .data$ADI_NATRANK,
        color = .data$Color
    )) +
    ggplot2::geom_point() +
    ggplot2::scale_x_continuous(
        name = "State Percentile of ADI",
        labels = scales::label_percent(accuracy = 1,
                                       scale = 10)
    ) +
    national_rank_axis +
    ggplot2::scale_color_manual(
        name = NULL,
        values = rlang::set_names(
            c(
                ggplot2::alpha(hiRx::higher_ex_cols("Gray"),
                               alpha = 1/128),
                ggplot2::alpha("red",
                               alpha = 10/128),
                hiRx::higher_ex_cols("Navy")
            ),
            nm = c(.WHOLE_US, .STATE_NAME, .COUNTY_NAME)
        )
    ) +
    generic_settings

```

### A Sanity Check Scatterplot of County Data

Before we draw the map, lets make sure things seem OK by plotting land area by ADI.
It won't look like there's a problem, now, but prepping for this image helped me refine a lot of the wrangling code above.

```{r area_by_adi_in_county}

COUNTY %>%
    ggplot2::ggplot(ggplot2::aes(
        x = .data$ALAND,
        y = .data$ADI_NATRANK,
        color = .data$ADI_NATRANK
    )) +
    ggplot2::geom_point(
        show.legend = FALSE
    ) +
    ggplot2::scale_x_continuous(
        name = "Block Group Land Area (sq. km.)",
        labels = scales::label_number(accuracy = 0.1,
                                      scale = 1e-6),
        trans = "log10"
    ) +
    national_rank_axis +
    rlang::exec(
        ggplot2::scale_color_gradient2,
        name = NULL,
        !!!national_rank_gradient_settings
    ) +
    generic_settings

```

### Draw the Chloropleth Map

As you may have noticed above, we like to color our maps with a yellow-gray-blue gradient.
Yellow, like the abdomen of many a wasp, indicates "Watch Out!".
Gray values are intermediate.
A soothing blue indicates areas where the system indicator suggests prosperity.
You can really see the benefit of re-using settings in this code block.

```{r map_adi_in_county}

COUNTY %>%
    ggplot2::ggplot(ggplot2::aes(
        fill = .data$ADI_NATRANK
    )) +
    ggplot2::geom_sf() +
    rlang::exec(
        ggplot2::scale_fill_gradient2,
        name = "National Percentile of ADI",
        !!!national_rank_gradient_settings
    ) +
    generic_settings

```

[neighborhood-atlas]: https://www.neighborhoodatlas.medicine.wisc.edu/
[tigerline]: https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2020&layergroup=Block+Groups
[nhgis]: https://www.nhgis.org/
[na-explanation]: https://www.neighborhoodatlas.medicine.wisc.edu/changelog
[census-fips-explanation]: https://www.census.gov/programs-surveys/geography/technical-documentation/records-layout/2020-census-block-record-layout.html
