---
title: "Style Reference"
author: "Higher Expectations for Racine County"
date: "`r format(Sys.time(), '%d %B, %Y')`"
# output:
#    word_document:
#        reference_docx: "HE style reference.docx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Word Style Reference Document}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hiRx)
```

# Style Reference

[RStudio](https://rstudio.com) uses the [`bookdown`](https://bookdown.org) to render Word documents.
The developers of `bookdown` decided that the most effective way to manage styling for Word documents was for the user to create a reference document with customized styles.
This document is an Rmarkdown progenitor for just such a style reference document.

## Second-level Heading

### Third-level Heading

#### Fourth-level Heading

```{r a-figure, fig.cap = "A figure caption"}

ggplot2::ggplot() +
    ggplot2::geom_text(ggplot2::aes(x = 1,
                                    y = 1,
                                    label = "Hello, World")) +
    ggplot2::theme_void() +
    ggplot2::theme(text = ggplot2::element_text(family = "Myriad Pro"))

```

A table

```{r a-table}

knitr::kable(tibble::tribble(
    ~ Foo, ~ Bar, ~ Baz,
    1, "a", TRUE,
    2, "b", FALSE,
    3, "c", NA
))

```
